---
title: "Reproducible Research Project 1"
author: "Kent Lanclos"
date: "April 7, 2018"
output: html_document
---
Load and preprocess the data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(plyr)
library(lattice)
# Load and preprocess the data
##########
mydata <- read.csv("activity.csv")
mydata$day <- weekdays(as.Date(mydata$date))
mydata$DateTime<- as.POSIXct(mydata$date, format="%Y-%m-%d")
```
Calculate the total number of steps taken per day
Create histogram of number steps per day

```{r }
numsteps <- aggregate(mydata$steps ~ mydata$date, FUN=sum, )
colnames(numsteps)<- c("Date", "Steps")
hist(numsteps$Steps, breaks=5, xlab="Steps", main = "Number of Steps per Day")
```

Calculate the mean and median of the total number of steps taken per day


```{r }
as.integer(mean(numsteps$Steps))
as.integer(median(numsteps$Steps))
```

Create average number of steps per interval
```{r}
nonadata <- mydata[!is.na(mydata$steps),]
stepsint <- ddply(nonadata, .(interval), summarize, Avg = mean(steps))
```

Create line plot of average number of steps per interval
```{r}
p <- ggplot(stepsint, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```
Maximum steps by interval
Which interval contains the maximum average number of steps

```{r}
stepsmax <- max(stepsint$Avg)
stepsint[stepsint$Avg==stepsmax,1]

```
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
nrow(mydata[is.na(mydata$steps),])
```
Substitute for NAs with the average 5-minute interval for day of the week
```{r}
avgnona <- ddply(nonadata, .(interval, day), summarize, Avg = mean(steps))
nadata <- mydata[is.na(mydata$steps),]
newdata <- merge(nadata, avgnona, by=c("interval", "day"))
```
Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
newdata2 <- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")
mrgdata <- rbind(nonadata, newdata2)
```
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Panel 1 has mean and median, Panel 2 has histogram. 
Mean is 10821, median is 11015. Values are higher than first estimates that included missing values.
```{r}
numsteps2 <- aggregate(mrgdata$steps ~ mrgdata$date, FUN=sum, )
colnames(numsteps2)<- c("Date", "Steps")
as.integer(mean(numsteps2$Steps))
as.integer(median(numsteps2$Steps))
hist(numsteps2$Steps, breaks=5, xlab="Steps", main = "Steps per Day with NAs Fixed", col="Black")
hist(numsteps$Steps, breaks=5, xlab="Steps", main = "Steps per Day with NAs Fixed", col="Grey", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey") )
```
Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
mrgdata$Category <- ifelse(mrgdata$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
```
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
stepsint2 <- ddply(mrgdata, .(interval, Category), summarize, Avg = mean(steps))
xyplot(Avg~interval|Category, data=stepsint2, type="l",  layout = c(1,2), main="Average Steps per Interval Based on Type of Day",ylab="Average Number of Steps", xlab="Interval")
```